---
title: "New York Flights of 2019"
date: "12/6/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
##Loading Libraries (these might need installation)


#importing libraries  
```{r libraries}
# #installing packages
# install.packages("RSQLite")
# install.packages(c('devtools','curl'))
# install.packages("dplyr")
# install.packages("ggplot2")
# install.packages("glmnet")
# install.packages("caret")

library(dplyr)
library(RSQLite)
library(vctrs)
library(ggplot2)
library(glmnet)
library(gplots)
library(ROCR)
library(caret)
```
## Data Setup

Steps to bring to set up the data are:  
1) Load the SQLite database  (For this to work, please paste your ) 
2) Create a dataframe for each table  
3) Join the 3 lookup tables to the ny_flights table for origin and destination airports, city, and airline_name   
4) Getting to know the data more, by selecting relevant columns:  
Relevant columns are flight time, airline, origin/destination city/airport, scheduled/actual departure, scheduled/actual arrival (in case there are patterns like flights that are supposed to arrive really late rarely get delayed) and delay in minutes for both, scheduled/actual elapsed time, taxi in/out


```{r data}
#bringing in the data from sqlite database
con <- dbConnect(drv=RSQLite::SQLite(), dbname="/Users/Downloads/Take-Home_Materials/flights.sqlite")
tables <- dbListTables(con)


## create a data.frame for each table

for (table in tables) {
  print(paste("select * from",table))
  sendquery_character <- paste0(table, "<- dbSendQuery(con, 'select * from ", table, "')")
  eval(parse(text = sendquery_character)) 
  fetch_character <- paste0(table, "<- dbFetch(", table, ")")
  eval(parse(text = fetch_character))
  dbClearResult(dbListResults(con)[[1]])
  eval(parse(text = paste0("print(head(", table, "))")))
}

#checking that the above worked
x <- dbSendQuery(con, "select count(*) from ny_flights", n=-1 )
y<- dbFetch(x)

dim(ny_flights)

#bringing in the city, airport, and airline into the ny_flights data frame
#for origin and destination cities
data<-dplyr::inner_join(ny_flights, cities, by = c("origin_city_market_id"="id"))
data$origin_city<-data$city
data<-dplyr::inner_join(data, cities, by = c("dest_city_market_id"="id"))
data$dest_city<-data$city.y

#for origin and destination airports
data<-dplyr::inner_join(data, airports, by = c("origin_airport_id"="id"))
data$origin_airport<-data$airport_name
data<-dplyr::inner_join(data, airports, by = c("dest_airport_id"="id"))
data$dest_airport<-data$airport_name.y

#for airline
data<-dplyr::inner_join(data, airlines, by = c("op_carrier_airline_id"="id"))

#selecting the relevant columns  using this as get to know what's in the data time
#relevant columns are flight time, airline, origin/destination city/airport, scheduled/actual departure
#scheduled/actual arrival (in case there are patterns like flights that are supposed to arrive really late rarely get delayed)
#and delay in minutes for both, scheduled/actual elapsed time, taxi in/out

names(data)
data2 <-dplyr::select(data, 
                    fl_date,
                    op_carrier_fl_num,
                    airline_name,
                    origin_airport,
                    origin_city,
                    dest_airport,
                    dest_city,
                    crs_dep_time,
                    dep_time,
                    dep_delay,
                    arr_delay, 
                    cancelled,
                    diverted,
                    crs_arr_time,
                    arr_time,
                    arr_delay,
                    crs_elapsed_time,
                    actual_elapsed_time,
                    taxi_in,
                    taxi_out
                    )
```

## Part 1: Exploring a route

Choose a route between two airports that you're interested in:  
I am interested in the NY-Miami route because I am constantly wishing I was in Florida on the beach where it's warm.

Some example questions you might investigate:

  - How busy is this route? Are there more flights on some days than others?
  - Which airlines provide the most flights on this route?
  - What do delays on this route look like?
  - Are there strange-looking days on this route with fewer-than-normal flights or
    longer-than-normal delays?

  - How busy is this route?
  
This route is very busy, in fact it is the second busiest route in the dataset.

```{r popular_route}
data2%>%
  filter(origin_city=='New York City, NY (Metropolitan Area)')%>%
  select(dest_city)%>%
  group_by(dest_city)%>%
  tally()%>%
  arrange(desc(n))
```

Miami is the second most popular route, out of NY! (This is looking by city, not airport)

Okay so what airports are in  Miami (I have never actually been), so I can pick a route, and which one has the most flights to and from NY?

```{r MiamiAirports}

data2%>%
  filter((origin_city=='New York City, NY (Metropolitan Area)' |
                  dest_city=='New York City, NY (Metropolitan Area)'),
         (origin_city=='Miami, FL (Metropolitan Area)' |
            dest_city=='Miami, FL (Metropolitan Area)'))%>%
  select(origin_airport,dest_airport)%>%
  group_by(origin_airport,dest_airport)%>%
  tally()%>%
  arrange(desc(n))

```


Looks like the LaGuardia-Miami International are the most popular.

  - Are there more flights on some days than others?   
  
We can see from the visualization that Saturday has less flights (1678 flights) than all the other days of the week , the rest of which are very similar averaging 1862 flights per day.

```{r nyfl}
#Limiting the data to flights from and to LaGuardia/Miami International:
nyfl<- dplyr::filter(data2,dest_airport=='LaGuardia'| origin_airport=='LaGuardia',
                     dest_airport=='Miami International'| origin_airport=='Miami International')

#creating a day of the week column
nyfl$day<-weekdays(as.Date(nyfl$fl_date,'%Y-%m-%d'))
nyfl$day<- factor(nyfl$day, levels=c("Monday","Tuesday", "Wednesday", "Thursday", "Friday", "Saturday","Sunday"))

ggplot(data=(dplyr::count(nyfl,day)), aes(x = day, y=n, fill=day)) +
  geom_bar(stat = "identity")+ggtitle(label = "Flights by Day of the Week")+
  theme_minimal()+theme(plot.title = element_text(hjust = 0.5, lineheight = 0.8, face = "bold"))+
  xlab("Day of the Week")+
  ylab("Number of Flights")

#What's the average number of flights? With and  without Saturday
y<-dplyr::count(nyfl,day)
mean(y$n)
y2<- dplyr::filter(y, day!='Saturday')
mean(y2$n)
```

   - Which airlines provide the most flights on this route?
American Airlines Inc. provides the most flights between LaGuardia and Miami International, with 64% of flights

```{r airlines}
nyfl%>%
  group_by(airline_name)%>%
  tally()%>%
  arrange(desc(n))%>%
  mutate(freq = n / sum(n))
``` 

  - What do delays on this route look like?  
  
When looking at delays, I see there  are 188 cancelations on this route.
I'm assuming anything above 0 is a delay, and below 0 is early, and not doing any transformations to account for a range where a flight is still considered on time. Most flights on this route depart early surprisingly! 56% of flights on this route depart early
and 37% depart late. Unsurprisingly, the same goes for arrivals, where 60% arrive early, and 36% arrive late.
Restating my assumption that there is probably a range where a flight could be considered on time.

```{r delays}
# Departure delays
hist(nyfl$dep_delay)
hist(nyfl$arr_delay)

nyfl$delay_cat[nyfl$dep_delay==0] <- "ontime"
nyfl$delay_cat[nyfl$dep_delay>0] <- "delay"
nyfl$delay_cat[nyfl$dep_delay<0] <- "early"

nyfl%>%
  group_by(delay_cat)%>%
  tally()%>%
  arrange(desc(n))%>%
  mutate(freq = n / sum(n))

# Arrival delays
nyfl$arr_cat[nyfl$arr_delay==0] <- "ontime"
nyfl$arr_cat[nyfl$arr_delay>0] <- "delay"
nyfl$arr_cat[nyfl$arr_delay<0] <- "early"

nyfl%>%
  group_by(arr_cat)%>%
  tally()%>%
  arrange(desc(n))%>%
  mutate(freq = n / sum(n))
```

  - Are there strange-looking days on this route with fewer-than-normal flights or longer-than-normal delays?  
  
Thir route has the highest frequency of flights per day during the holidays (December 22nd to the end of the year), which is not surprising. The median number of flights is 34 flights per day, and from the histogram below, it is clear that days with less than 30 flights have a lower number of flights.  
March has the most days with less than 30 flights, March 21st-March 31st, to be exact have a particularly low number of flights.
This could be a result of Spring Break seasonality. June has the highest number of flights in this route, where every day has >=34 flights.  
As for delays, I hypothesized that winter days would have the highest delays, specifically
out of LaGuardia, but turns out summer months (June, July, August), have by far the highest number of delays!
This could be a result of the high volume of daily flights on this route.


```{r strange_looking_days}
#to look at flight frequency by day 
nflights <- dplyr::count(nyfl,fl_date)

# Plot the chart with special scale lables
nflights<- nyfl%>%
  group_by(fl_date)%>%
  tally()%>%
  arrange(desc(n))

qplot(nflights$n,
      geom="histogram",
      binwidth =1,  
      main = "             Histogram of Number of Flights by Day", 
      xlab = "N Flights Per Day",  
      ylab = "Frequency",  
      fill=I("blue"), 
      col=I("red"), 
      alpha=I(.2),
      xlim=c(20,40),
      ylim=c(0,120)
      )

#to look at days with a low number of flights
lown<-dplyr::filter(nflights, n<30)
#creating a month column
lown$month<-months(as.POSIXct(lown$fl_date, format="%Y-%m-%d"))
lown$month<- factor(lown$month, levels=c("March","April", "May", "June", "July", "August","September","October","November"))

# by month
nflights$month<-months(as.POSIXct(nflights$fl_date, format="%Y-%m-%d"))
nflights$month<- factor(nflights$month, levels=c("January","February","March","April", "May", "June", "July", "August","September","October","November","December"))

nflights%>%
  group_by(month)%>%
  summarise(min=min(n))

#plotting months to vizualize
ggplot(data=(dplyr::count(lown,month)), aes(x = month, y=n, fill=month)) +
  geom_bar(stat = "identity")+
  ggtitle(label = "N Days With Less Than 30 Flights by Month")+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  theme(plot.title = element_text(hjust = 0.5, lineheight = 0.8, face = "bold"))+
  xlab("Month")+
  ylab("Number of Days with less than 30 flights")

dplyr::filter(lown, month=='March')


#plotting months to vizualize
nyfl$month<-months(as.POSIXct(nyfl$fl_date, format="%Y-%m-%d"))
nyfl$month<- factor(nyfl$month, levels=c("January","February","March","April", "May", "June", "July", "August","September","October","November","December"))

nflights<- nyfl%>%
  group_by(month)%>%
  tally()%>%
  arrange(desc(n))

ggplot(data=nflights, aes(x = month, y=n, fill=month)) +
  geom_bar(stat = "identity")+
  ggtitle(label = "Flights by Month")+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  theme(plot.title = element_text(hjust = 0.5, lineheight = 0.8, face = "bold"))+
  xlab("Month")+
  ylab("N Flights")

#any standout delays on specific days? checking to see if days are different when NY is the origin 
# Hypothesized that winter has highest delays, doesn't seem like that's true.
sumdelaysbymonth<- 
  nyfl %>%
  group_by(month) %>%
  filter(is.na(dep_delay)==FALSE)%>%
  summarise(sum = sum(dep_delay))%>%
  arrange(desc(sum))

ggplot(data=sumdelaysbymonth, aes(x = month, y=sum, fill=month)) +
  geom_bar(stat = "identity")+
  ggtitle(label = "Net Delay in Minutes by Month")+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  theme(plot.title = element_text(hjust = 0.5, lineheight = 0.8, face = "bold"))+
  xlab("Month")+
  ylab("Net Delay in Minutes")
```

##Part 2: Comparing Airports

Create some summary data and visualizations comparing the three NYC-area airports.
Of particular interest are departure delays from the airports. Are there differences in
the probability of a late (say, more than 30 min.) departure? How does time spent on the
runway compare amongst the airports?

Assuming that the three NYC-area airports are LaGuardia, JFK, Newark, and limiting the data
to flights that originate from those airports.

First thing to notice is how few flights are out of Newark. There appear to be 2877 flights out of Newark for the whole year which is only ~8 flights a day which sounds incorrect. Assuming this data is correct, La Guardia has the highest share of Flights out of the three airports.

Regarding to delays, ironically, Newark has the highest percent of delays at 23% of flights out of Newark having delays. LaGuardia has the second highest percent of delays at 16%, followed by JFK at 13%.
In addition to Newark having the highest probability of having a delay, cancelations also are twice as likely at Newark (6.5%) than they are at Laguardia, and less so at JFK.  

As for time spent on the runway (using taxi_out for this metric), the three airports have
similar taxi time, with Newark's slightly lower (24.8 minutes on average). JFK and LaGuardia
have an average of 27.6 and 27 per flight minutes respectively.

```{r comparing_airports}

nyairports<-dplyr::filter(data2,origin_city=='New York City, NY (Metropolitan Area)')
dplyr::count(nyairports, origin_airport)

#getting the correct airports 
nyairports<-dplyr::filter(data2,origin_airport=='John F. Kennedy International'|origin_airport=='LaGuardia'|origin_airport=='Newark Liberty International')
nyairports$delay_cat<- ifelse(nyairports$dep_delay>=30,1,0)
nyairports$delay_cat[nyairports$cancelled==0 & is.na(nyairports$dep_delay)==TRUE] <- 0

dplyr::count(nyairports, origin_airport, delay_cat)

nairports<- as.data.frame(nyairports%>%
  group_by(origin_airport)%>%
  tally()%>%
  arrange(desc(n))%>%
  mutate(freq = n / sum(n)))

ggplot(nairports, aes(x=origin_airport,y = n, label = n, fill=origin_airport)) +
  geom_col() +
  geom_text(nudge_y = 100) + 
  ggtitle(label = "Number of Flights by NY Airport")+
  theme_minimal()+
  theme(plot.title = element_text(hjust = 0.5, lineheight = 0.8, face = "bold"))+
  theme(legend.position = "none")+
  xlab("Airport")+
  ylab("N Flights")
```

 - Are there differences in the probability of a late (say, more than 30 min.) departure?  
 
Looking at the data, seeing that there are some NA dep_delay, but with crs_dep_time==dep_time. i I will assume there is no delay there.  Newark has the highest % of delays, and cancelations.
Similar for taxi time, Newark has the lowest.

```{r probs_delay}

nyairports%>%
  filter(cancelled==0)%>%
  count(origin_airport,delay_cat)


delays<-
  dplyr::inner_join(nyairports%>%
  filter(cancelled==0)%>%
  count(origin_airport,delay_cat),
  (nyairports%>%
     filter(cancelled==0)%>%
     count(origin_airport)), by='origin_airport')

delays<-as.data.frame(dplyr::rename(delays, n=n.x, total=n.y))

#delay % of flights
delays %>% 
  group_by(origin_airport) %>%
  filter(delay_cat==1)%>%
  summarise(n/total)

delays$percent_delay<- round((delays$n*100)/delays$total)
  
delays$delay_cat<-ifelse(delays$delay_cat==1,"delayed","on time")

#creating the correct position for the %s
library(plyr)
delayplot <- ddply(delays, .(origin_airport),
                     transform, pos = cumsum(percent_delay) - (0.5 * percent_delay))
detach(package:plyr)

ggplot() + geom_bar(aes(y = percent_delay, x = origin_airport, fill = delay_cat), data = delayplot,
                          stat="identity") +

  geom_text(data=delayplot, aes(x = origin_airport, y = pos,
                                           label = paste0(percent_delay,"%")), size=4)



#cancelations
cancelations<-
  dplyr::inner_join(nyairports%>%
                      count(origin_airport,cancelled),
                    (nyairports%>%
                       count(origin_airport)), by='origin_airport')

cancelations<-as.data.frame(dplyr::rename(cancelations, n=n.x, total=n.y))


cancelations %>% 
  group_by(origin_airport) %>%
  filter(cancelled==1)%>%
  summarise(n/total)

#time on the runway (taxi_out)

time_taxi<-  as.data.frame(
  dplyr::inner_join(nyairports %>% 
  group_by(origin_airport) %>%
  filter(is.na(taxi_out)==FALSE)%>%
  summarise(total_taxitime=sum(taxi_out)),
  nyairports %>% 
    group_by(origin_airport) %>%
    filter(is.na(taxi_out)==FALSE)%>%
    tally, by='origin_airport'))

  
time_taxi %>% 
  group_by(origin_airport) %>%
  summarise(total_taxitime/n)
```

##Part 3  

1. Build a simple model that estimates the probability that a given flight will have a late
(greater than 30 min.) departure or arrival.

- You can interpret the term "model" loosely: Any method that, given some inputs known prior to a flight,
will produce an estimated probability that the flight will be
delayed or arrive late. You should also be able to use the model to show the uncertainty
around the estimate.

```{r model_create}
#setting up the basefile, no cancelled flights,
#dependent variable is delay: 1 or 0

#looking at nulls, looks good mostly
bf<-dplyr::filter(data2, cancelled==0)

apply(bf,2,function(x){sum(is.na(x))})


#creating a day of the week column, month
bf$day<-weekdays(as.Date(bf$fl_date,'%Y-%m-%d'))
bf$month<-months(as.POSIXct(bf$fl_date, format="%Y-%m-%d"))

#creating time of day column
bf$timeofday[bf$crs_dep_time<1200] <-'a-Morning'
bf$timeofday[bf$crs_dep_time>=1200 & bf$crs_dep_time<1800] <-'b-Afternoon'
bf$timeofday[bf$crs_dep_time>=1800] <-'c-Night'

#checking flights length distribution
hist(bf$crs_elapsed_time)
# ok, most flights are under 200 minutes. setting long>200
bf$long_flight<- ifelse(bf$crs_elapsed_time>200,1,0)

#creating the dependent variable
bf$delay_cat<- ifelse(bf$dep_delay>=30,1,0)
bf$delay_cat[bf$crs_dep_time==bf$dep_time] <- 0
bf<-dplyr::filter(bf, is.na(delay_cat)==FALSE)


#using stepwise selection using a logistic regression first to see what features are significant
# in prediciting delays

# logit <- glm(delay_cat ~ origin_city, family = "binomial", data=bf)

# summary(logit)

  
#looks like saturday, sunday, Wednesday have high/significant coefs
#Frontier, SkyWest,  ExpressJet  have very high coefs
#Night has high coefs 
  
bf$airline_highdelay<- ifelse(bf$airline_name %in%  c('ExpressJet Airlines LLC','SkyWest Airlines Inc.','Frontier Airlines Inc.'),1,0)
bf$weekend<- ifelse(bf$day %in%  c('Saturday','Sunday'),1,0)

# bf <- dplyr::select(bf
#                    , long_flight
#                    , timeofday
#                    , month
#                    # , weekend
#                    , day
#                    , airline_highdelay
#                    , delay_cat)

#creating training/test set
bf$holdout = runif(nrow(bf))
train <- bf[bf$holdout < .7,]
test <- bf[bf$holdout >= .7,]

formula = as.formula(delay_cat~  long_flight + timeofday + month + weekend + airline_highdelay)

#trying interactions
# formula2 = as.formula(delay_cat~  (long_flight + timeofday + month + weekend + airline_highdelay)^2)

#tried formula with month*weekend, but it made AUC worse

mm <- model.matrix(formula,data=train)
mmtest <- model.matrix(formula, data=test)
mmall <- model.matrix(formula, data=bf)

plot(glmnet(mmall[ , -1], bf$delay_cat))

#train a lasso model for lamda.1se, alpha=1
train.cvmn <- cv.glmnet(mm[ , -1], train$delay_cat, family="binomial", standardize=FALSE, type.measure="auc", nfold=10,alpha=1)
cvmnf <- glmnet(mm[ , -1], train$delay_cat,family="binomial",alpha=1,standardize = FALSE, lambda = train.cvmn$lambda.1se)
plot(train.cvmn) 

#looking at the most important features 
coef <- coef(cvmnf, s=cvmnf$lambda.min)[order(abs(coef(cvmnf, s=cvmnf$lambda.min)), decreasing = TRUE)][1:15]
names <- rownames(coef(cvmnf, s=cvmnf$lambda.min))[order(abs(coef(cvmnf, s=cvmnf$lambda.min)), decreasing = TRUE)][1:15]
for (iii in 1:15){
  cat(names[[iii]],"\t",coef[[iii]],"\n")
} 


calc_AUC <- function(obs, pred, plotROC=FALSE) {
  roc.pred<-prediction(pred, obs)
  if (plotROC) {
    roc.perf <- ROCR::performance(roc.pred, measure="tpr", x.measure="fpr")
    plot(roc.perf)
    abline(a=0,b=1)
  }
  return(ROCR::performance(roc.pred, "auc")@y.values[[1]])
}

#scoring 
train$delay_score <- as.numeric(predict(cvmnf, newx=mm[ , -1], type="response", s=cvmnf$lambda.1se))

#AUC
calc_AUC(train$delay_cat, train$delay_score, "TRUE")

hist(train$delay_score)

#out of sample
test$delay_score <- as.numeric(predict(cvmnf, newx=mmtest[ , -1], type="response", s=cvmnf$lambda.1se))
calc_AUC(test$delay_cat, test$delay_score , "TRUE")
hist(test$delay_score)

test$delay_pred<-ifelse(test$delay_score>=0.17,1,0)
confusionMatrix( as.factor(test$delay_pred),as.factor(test$delay_cat), positive="1")

```

2. Discuss the results of your fitted model. Here are a few things to describe:

  - The performance of your model---either its in-sample fit, or out-of-sample prediction
accuracy (or both).
- For a few interesting example flights or routes, use the model to estimate probability of
a late departure or arrival and its uncertainty. Describe the results in plain English.
- If your model is suited to estimating factors that affect delays
(that is, it has interpretable coefficients), describe the estimated effects.

My Lasso model has an out of sample AUC of 65, in sample 66.
Using this model, a test set flight on Frontier Airlines, from LaGuardia to Atlanta, flying in June, on a Monday in the afternoon, has a 26% probability of being delayed. Using the threshold 17%, this flight will be predicted to be delayed.
Given the AUC of 65 (out of sample), we understand that there is a 40% chance this is a False Positive (will not actually be delayed), and a 60% chance this is a True positive (will in fact be delayed).

According to this model, features that are likely to predict a delayed flight are:
A night time flight, being on one of these airlines: ExpressJet Airlines LLC','SkyWest Airlines Inc.','Frontier Airlines Inc.', flying in July, August, and  to a lesser degree, January and  June.

Features that are less likely to predict a delayed flight:  
Being a weekend flight (Saturday and Sunday), flying in  March, September, and being a long flight.

However, just noting that with Lasso, it is not always very clear that coeficients represent
the magnitude of the effect of features.


3. Describe anything you feel might be missing from your model,
or alternative approaches you might try with more time or data.
If you think additional data would be helpful, describe what you would want.

Given more time, I would have focused a lot more on the seasonality element, and did more work
to find origin/destination combinations that have delays. For example, if a route has many flights a day, that couldmake flights more likely to be delayed, flights originating or heading to particular destinations could be more likely to be delayed.
I probably would have considered doing a stepwise model, instead of throwing most things and the kitchen sink in a lasso, as well as an xgboost model. I might have also used Facebook's Prophet package because it is good with seasonality. I would have also looked more into problems with the model like multicollinearity and outliers in the additional data I would have included.
